\section{Non-preemptive Interruptions, Server Adjustments}
\label{sec:non-preempt-interr}


\opt{solutionfiles}{
\subsection*{Theory and Exercises}
\Opensolutionfile{hint}
\Opensolutionfile{ans}
}

%See section 1.11 of Zijm's book for the theory.


In~\cref{sec:setups-batch-proc} we studied the effect of setup times between job batches with a fixed size~$B$.
%In other words, there are no setups between any two jobs, setups are \emph{planned} between $B$ jobs.
However, other types of interruptions can occur, such as a machine requiring random adjustments that can occur between any two jobs.
This type of outages is \recall{non-preemptive} as the outages do not interrupt the processing of a job in service.
In this section we develop a simple model to understand the impact of such outages on job sojourn times; we use the same notation as in~\cref{sec:setups-batch-proc} and follow the same line of reasoning.

We assume that adjustments $\{R_i\}$ occur geometrically distributed between any two jobs with a mean of $B$ jobs between any two adjustments.
Consequently, the probability of an outage between any two jobs is $p=1/B$.
Observe that geometrically distributed random variables satisfy the memoryless property in discrete time.
Hence, our assumption implies that the occurrence of an adjustment between jobs $i$ and $i+1$ has no effect on the probability that an adjustment is necessary between jobs $i+1$ and $i+2$.

With the model, we can obtain quantitative insights into the effects of reducing adjustment times or the variability of these adjustments times.
For instance, we might decide to do fewer adjustments, so that $p$ decreases, but at the expense of larger average outage times.
Thus, without needing to actually do the experiments in real life, we can evaluate the influence of such decisions.

Contrary to the batch processing case of~\cref{sec:setups-batch-proc}, we now only need to find the mean and variance of the effective processing times.
From~\cref{ex:l-256} and \cref{,ex:l-257} it follows that
\begin{exercise}\clabel{ex:l-256}
Show that the average effective processing time is given by
\begin{equation} \label{eq:88}
 \E{S} = \E{S_0} + \frac{\E R}B
\end{equation}

\begin{hint}
  If there is no interruption, the service time is $S_0$.
  However, if there is an interruption, the service of a job is $R + S_0$, since we add the outage time to the service time of the job.
  Relate this insight to~\cref{ex:49}.
  In that exercise, take $p=(B-1)/B$ as the probability of a short job of size $\E{S_0}$ and $q=1/B$ as the probability of a long job of $\E{S_0} + \E{R}$.
\end{hint}
\begin{solution}
 \begin{equation*}
 \E{S} = (1-p)\E{S_0} + p (\E{R} + \E{S_0}) = \E{S_0} \frac{B-1}B + (\E{S_0} + \E R) \frac 1B,
 \end{equation*}
since $p=1/B$. 
\end{solution}
\end{exercise}


%The next step is to compute $\E{S^2}$. 
\begin{extra}\clabel{ex:78}
 Show that
 \begin{equation*}
 \E{S^2} = \E{S_0^2} + 2\frac{\E{S_0}\E{R}} B + \frac{\E{R^2}}B.
 \end{equation*}
\begin{hint}
  Again, use~\cref{ex:49}.
\end{hint}
\begin{solution}
 \begin{align*}
 \E{S^2} 
&= (1-p) \E{S_0^2} + p \E{(S_0+R)^2}\\
&= (1-p) \E{S_0^2} + p \E{S_0^2} + 2 p \E{S_0} \E R + p \E{R^2}. 
 \end{align*}
 Simplify and substitute $p=1/B$. 
\end{solution}
\end{extra}

\begin{exercise}\clabel{ex:l-257}
Derive 
\begin{equation}\label{eq:89} 
 \V{S} = \V{S_0} + \frac{\V{R}} B + (B-1)\left(\frac{\E R}{B}\right)^2.
\end{equation}
\begin{hint}
 First compute $\E{S^2}$. See~\cref{ex:78}.
\end{hint}
\begin{solution}
 \begin{equation*}
 \begin{split}
\V{S} 
&=\E{S^2} - (\E{S})^2 \\
&= \E{S_0^2} + 2 \E{S_0} \E R\frac 1B + \E{R^2}\frac 1B \\
&\quad - (\E{S_0})^2 - 2\E{S_0}\E R \frac 1 B - (\E R)^2\frac1{B^2}\\
&= \V{S_0} + ((\E{R^2} - (\E R)^2)\frac 1 B + (\E R)^2\left(\frac 1B - \frac 1{B^2}\right).
 \end{split}
 \end{equation*}
\end{solution}
\end{exercise}


Thus, the effective server load including down-times is $\rho = \lambda \E{S}$, and we can compute the SCV of the effective job processing times.
With this, we have all elements to fill in the $G/G/1$ waiting time formula!


\begin{exercise}\clabel{ex:l-255}
 Jobs arrive as a Poisson process with rate $\lambda=9$ per working day.
 The machine works two $8$ hour shifts a day.
 Work not processed on a day is carried over to the next day.
 Job service times are 1.5 hours, on average, with standard deviation of $0.5$ hours.
 Outages occur on average between $30$ jobs.
The average duration of an outage  is of $5$ hours and has a standard deviation of $2$ hours.

 Compute the average waiting time in queue.
\begin{hint}
 Get the units right. First compute the load, and then compute the rest.
\end{hint}
\begin{solution}
 First we determine the load. 
 \begin{pyconsole}
B=30
ES0 = 1.5
labda = 9./(2*8) # arrival rate per hour
ER=5.
ESe=ES0+ER/B
ESe
rho = labda*ESe
rho
 \end{pyconsole}
So, at least the system is stable.

\begin{pyconsole}
VS0 = 0.5*0.5
VR = 2.*2.
VSe = VS0 + VR/B + (B-1)*(ER/B)**2
VSe
Ce2 = VSe/(ESe*ESe)
Ce2
\end{pyconsole}

And now we can fill in the waiting time formula.
\begin{pyconsole}
Ca2=1 # Poisson arrivals
EW = (Ca2+Ce2)/2 * rho/(1-rho) * ESe
EW 
\end{pyconsole}
\end{solution}
\end{exercise}




\opt{solutionfiles}{
\Closesolutionfile{hint}
\Closesolutionfile{ans}
\subsection*{Hints}
\input{hint}
\subsection*{Solutions}
\input{ans}
}

%\clearpage

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../companion"
%%% End:
