\section{Rate, Stability and Load}
\label{sec:rate-stability}



In this section, we develop a number of essential concepts to analyze queueing systems: the arrival, service and departure rate.
With these we define the load, which is, arguably, the most important performance measure of a queueing system to check.
%We will use these concepts throughout the remainder of the book.

\opt{solutionfiles}{
\subsection*{Theory and Exercises}
\Opensolutionfile{hint}
\Opensolutionfile{ans}
}


We first formalize the arrival rate and departure rate in terms of the arrival and departure processes $\{A(t)\}$ and $\{D(t)\}$; recall that these are \emph{counting processes}.
The \recall{arrival rate} is the long-run average number of jobs that arrive per unit time along a sample path, i.e.,
\begin{equation}
 \label{eq:3}
 \lambda = \lim_{t\to\infty} \frac{A(t)}t.
\end{equation}
We remark in passing that this limit does not necessarily exist if $A(t)$ is some pathological function.
If, however, the inter-arrival times $\{X_k\}$ are the basic data, and $\{X_k\}$ are i.i.d.
and distributed as a generic random variable $X$ with finite mean $\E{X}$, we can construct $\{A_k\}$ and $\{A(t)\}$ as described in~\cref{sec:constr-gg1-queu}, and then the strong law of large numbers guarantees that the above limit exists.\footnote{In fact, $A(t)/t \to \lambda$ with probability one.}

\begin{extra}
  Can you make an arrival process such that $A(t)/t$ does not have a limit?
\begin{hint}
  As a start, the function $\sin(t)$ does not have a limit as $t\to\infty$.
  However, the time-average $\sin(t)/t \to 0$.
  Now you need to make some function whose time-average does not converge, hence it should grow fast, or fluctuate wilder and wilder.
\end{hint}
\begin{solution}
 If $A(t) = 3 t^2$, then clearly $A(t)/t = 3t$. This does not
 converge to a limit. 

 Another example, let the arrival rate $\lambda(t)$ be given as follows:
 \begin{equation*}
 \lambda(t) = 
 \begin{cases}
 1 & \text{if } 2^{2k} \leq t < 2^{2k+1} \\
 0 & \text{if } 2^{2k+1} \leq t < 2^{2(k+1)},
 \end{cases}
 \end{equation*}
 for $k=0,1,2,\ldots$.
 Let $A(t) = \lambda(t) t$.
 Then $A(t)/t$ does not have limit.
 Of course, these examples are quite pathological, and are not representable for `real life cases'.
 (Although this is also quite vague.
 What, then, is a real life case?)

 For the mathematically interested, we seek a function whose Ces\`aro limit does not exist.
\end{solution}
\end{extra}

Likewise, define the \recall{departure rate} as
\begin{equation}\label{eq:28}
 \delta = \lim_{t\to\infty} \frac{D(t)}t.
\end{equation}
Observe that, if the system is empty at time $0$, the number of departures must be smaller than or equal to the number of arrivals, i.e., $D(t) \leq A(t)$ for all $t$.
Therefore,
\begin{equation}\label{eq:26}
\delta = \lim_{t\to\infty} \frac{D(t)}t \leq \lim_{t\to\infty} \frac{A(t)}t = \lambda.
\end{equation}
It is evident that when $\lambda > \delta$, the system length process $L(t) \to \infty$ as $t\to \infty$.
We therefore call a system \recall{rate-stable} if
\begin{equation*}
 \lambda = \delta.
\end{equation*}
In  words: the system is rate-stable whenever jobs leave the system just as fast as they arrive in the long run.


\begin{exercise}\clabel{ex:98}
If the system starts empty, then we know that the number $L(t)$ in the system at time $t$ is equal to $A(t) - D(t)$.
Show that the system is rate-stable  if $L(t)$ remains finite, or, more generally, $L(t)/t \to 0$ as $t\to\infty$.
\begin{solution}
Since $L(t) = A(t) - D(t)$, 
\begin{equation*}
 \lambda = \lim_{t \to \infty} \frac{A(t)}t = \lim_{t \to \infty} \frac{D(t)+L(t)}t = \lim_{t \to \infty} \frac{D(t)}t + \lim_{t \to \infty} \frac{L(t)}t 
 = \delta. 
\end{equation*}
Hence, $\lambda=\delta$ when $L(t)/t\to0$.
\end{solution}
\end{exercise}


We next relate the arrival rate $\lambda$ to the expected inter-arrival time $\E X$.
Observe that at time $t=A_n$ precisely $n$ arrivals occurred.
Thus,
% by applying the definition of $A(t)$ at the epochs $A_n$,
we see that $A(A_n) = n$, and therefore
\begin{equation*}
 \frac{1}n\sum_{k=1}^n X_k = \frac{A_n}n = \frac{A_n}{A(A_n)}. 
\end{equation*}
But since $A_n\to\infty$ if $n\to\infty$, it follows from~\cref{eq:3} that the average inter-arrival time between two consecutive jobs is
\begin{equation}\label{eq:54}
 \E X = \lim_{n\to\infty} \frac{1}n\sum_{k=1}^n X_k = \lim_{n\to \infty} \frac{A_n}{A(A_n)} = \lim_{t\to\infty} \frac t{A(t)} = \frac 1 \lambda,
\end{equation}
where we take $t=A_n$ in the limit for $t\to\infty$.
In words, the arrival rate $\lambda$ is the \emph{inverse} of the expected inter-arrival time $\E X$.


Assume now that there is a single server.
Let $S_k$ be the required service time of the $k$th job to be served, so that  $U_n = \sum_{k=1}^n S_k$ becomes the total service time required by the first $n$ jobs.
With this, let $ U(t) = \max\{n: U_n \leq t\}$, so that we can define the \recall{service rate} or \recall{processing rate} as
\begin{equation*}
 \mu = \lim_{t\to\infty} \frac{U(t)}t.
\end{equation*}
Similar to the relation  $\E X= 1/\lambda$, we the relation
\begin{equation*}
 \E S = \lim_{n\to\infty} \frac 1 n \sum_{k=1}^n S_k = \lim_{n\to\infty} \frac{U_n}{n} = \lim_{n\to\infty} \frac{U_n}{U(U_n)} = \lim_{t\to\infty} \frac t{U(t)} = \frac 1 \mu.
\end{equation*}
Thus,  the service rate $\mu$ is the \emph{inverse} of the expected service time $\E S$.

It turns out that, when $\mu = \lambda$ and $\V{S} > 0$ or $\V{X} >0$, the queue length process behaves in a very peculiar way.
This follows from the fact that the random walk without drift, i.e., $\mu=\lambda$, has some unexpected behavior\footnote{See any good book probability theory.}, and we know from~\cref{sec:queu-proc-as} that queueing system and random walks are intimately related.
To avoid such problems, we henceforth (and implicitly) require that $\mu > \lambda$.
Observe also the evident fact that jobs cannot depart faster than they can be served, hence, $D(t) \leq U(t)$ for all~$t$, hence $\delta \leq \mu$.
Combining this with $\lambda < \mu$ and rate-stability, we see that $\delta < \mu$.


\begin{extra}
  Define the random variables $\{\tilde X_k,k=1,\ldots\}$ as $\tilde X_k = S_{k-1}-X_k$.
  For stability of the queueing process it is essential that $\tilde X_k$ has negative expectation, i.e., $\E{\tilde X_k} = \E{S_{k-1}-X_k} < 0$.
  What is the conceptual meaning of this inequality?
\begin{solution}
 That the average time customers spend in service is smaller
 than the average time between the arrival of two subsequent
 jobs. 
\end{solution}
\end{extra}

\begin{exercise}\clabel{ex:l-253}
Define $\tilde X_k = S_{k-1}-X_k$.
 Show that $\E{\tilde X_k} <0$ implies that $\lambda<\mu$. 
\begin{hint}
Remember that $\{X_k\}$ and $\{S_k\}$ are sequences of i.i.d. random variables. What are the implications for the expectations?
\end{hint}
\begin{solution}
 $0> \E{\tilde X_k} = \E {S_{k-1}-X_k} = \E{ S_{k-1}}- \E {X_k} = \E S - \E X$, where we use the fact that the $\{S_k\}$ and $\{X_k\}$ are i.i.d. sequences. Hence, 
 \begin{equation*}
 \E X > \E S \iff \frac 1{\E S} > \frac1{\E X} \iff \mu > \lambda.
 \end{equation*}

\end{solution}
\end{exercise}

Perhaps the most important performance measure is the concept of \recall{load}~$\rho$, which is defined as the rate at which jobs arrive times the average amount of work per job: $\rho = \lambda \E S$.
From the identities $\lambda^{-1} = \E X$ and $\mu^{-1} = \E S$, we get some further relations:
\begin{equation}
 \rho = \lambda \E S = \frac{\lambda}{\mu} = \frac{\E S}{\E X}.
\end{equation}
Observe that by assumption $\mu> \lambda$, hence $\rho < 1$.
The relation $\rho=\E S/ \E X < 1$ then tells us that the average time it takes to serve a job must be less than the average time between two consecutive arrivals, i.e., $\E S < \E X$.
Also, when $\lambda < \mu$, it is easy to check with simulation that when $L(0)$ is very large, $L(t) \approx L(0) - (\mu-\lambda)t$ until the system is empty, while if $\lambda>\mu$, we have that $l(t) \approx L(0) + (\lambda-\mu)t$. 


\begin{exercise}\clabel{ex:l-164}
 Consider a queueing system with $c$ servers with identical production rates $\mu$.
 What would be a reasonable stability criterion for this system?
\begin{hint}
What is the rate in, and what is the service capacity?
\end{hint}
\begin{solution}
 The criterion is that $c$ must be such that $\lambda < c\mu$.
 (Thus, we interpret the number of servers as a \emph{control}, i.e., a `thing' we can change, while we assume that $\lambda$ and $\mu$ cannot be easily changed.)
 To see this, we can take two different points of view.
 Imagine that the $c$ servers are replaced by one server that works $c$ times as fast.
 The service capacity of these two systems (i.e., the system with $c$ servers and the system with one fast server) is the same, i.e., $c\mu$, where $\mu$ is the rate of one server.
 For the system with the fast server, the load is defined as $\rho =\lambda/c\mu$, and for stability we require $\rho<1$.
 Another way to see it is to assume that the stream of jobs is split into $c$ smaller streams, each with arrival rate $\lambda/c$.
 In this case, applying the condition that $(\lambda/c )/\mu<1$ per server leads to the same condition that $\lambda/(c\mu) < 1$.
\end{solution}
\end{exercise}


\begin{extra}[Not obligatory]
 In~\cref{eq:54} we replaced the limit with respect to $n$ by a limit with respect to~$t$.
 Use the notation $A_{A(t)}$ to show that this is allowed.
 Show next that the function $t\to A(t)$ as defined by~\cref{eq:2} is right-continuous.
\begin{hint}
 Use that $A_{A(t)} \leq t < A_{A(t)+1}$.
 Divide by $A(t)$ and take suitable limits.
 BTW, such type of proof is used quite often to show that the existence of one limit implies, and is implied by, the existence of another type of limit.
\end{hint}
\begin{solution}
 This exercise is meant to provide some insight into what needs to be done to put everything on solid ground.


 Observing that $A_{A(t)}$ is the arrival time of the last job before time $t$ and that $A_{A(t)+1}$ is the arrival time of the first job after time $t$:
 \begin{equation*}
 A_{A(t)} \leq t < A_{A(t)+1} \Leftrightarrow 
 \frac{A_{A(t)}} {A(t)} \leq \frac{t}{A(t)} <\frac{A_{A(t)+1}}{A(t)} = \frac{A_{A(t)+1}}{A(t)+1}\frac{A(t)+1}{A(t)}.
 \end{equation*}
 Now $A(t)$ is a counting process such that $A(t)\to\infty$ as $t\to\infty$.
 Therefore, $\lim_{t\to\infty} A_{A(t)}/A(t) = \lim_{n\to\infty} A_n/n$.
 Moreover, it is evident that $\lim_{t\to\infty} A_{A(t)+1}/(A(t)+1) = \lim_{t\to\infty} A_{A(t)}/A(t)$, and that $(A(t)+1)/A(t)\to 1$ as $t\to\infty$.
 Thus it follows from the above inequalities that $\lim_{n\to\infty} A_n/n = \lim_{t\to\infty} t/A(t)$.
 
 For the right-continuity of $A(t)$, define $f(t) = \1{A_1 \leq t}$.
 Observe first that $f(t)$ is increasing, and $f(t)\in\{0,1\}$.
 Thus, if $f(t)=1$ then $f(u)=1$ for all $u\geq t$, and if $f(t)=0$ then $f(u) = 0$ for all $u\leq t$.

 You may skip the rest of the proof below, but the above is essential to memorize; make a plot of $f(t)$, in particular, the behavior around $A_1$ is important.

 We need to prove, for right-continuity, that $f(u)\to f(t) $ as $u\downarrow t$.
 When $f(t)=1$, $f(u)=1$ for any $u>1$, by the definition of $f(x)$.
 When $f(t)=0$ we have to do a bit more work.
 Formally, we have to prove that, for fixed $t$ and for all $\epsilon > 0$, there is a $\delta>0$ such that $u\in(t, t+\delta) \Rightarrow |f(u) -f(t)| < \epsilon$.
 (Note the differences with the regular definition of continuity.)
 Since, by assumption, $t$ is such that $f(t)=0$, and $f\in\{0,1\}$ we need to show that $f(u)=0$ for $u\in(t, t+\delta)$.
 Now, clearly, $f(t)=0$ only if $t < A_1$.
 But, then for any $u\in(t, A_1)$, we have that $f(u) = 0$.
 Thus, taking $\delta = A_1 - t$ suffices.

 The next step is to observe that $A(t)$ is a sum of right-continuous functions whose steps do not overlap since by assumption $0<A_1 < A_2 < \cdots$.
 As $A$ is (almost surely) a finite sum of bounded, increasing and right-continuous functions, it is also right-continuous.

 If you like, you can try to prove this last step too. 

 Hopefully this problem, and its solution, clarifies that even such small details require attention.
 If we want to make some progress with respect to developing some queueing theory, we have to skip most of the proofs and mathematical problems; we simply don't have enough time in this course to be concerned with all theorems and proofs.

 % For finite sums it is simple. Suppose that $f$ and $g$ are
 % right-continuous, then
 % \begin{equation*}
 % |f(u) + g(u) - f(t) - g(t)| \leq |f(u)-f(t)|+|g(u)-g(t)|.
 % \end{equation*}
 % Since both terms at the right-hand side can be made arbitrarily
 % small, the left-hand side can also be made as small as we
 % like. With this we can see that the function
 % $F_N(t) = \sum_{n=1}^N f_n(x)$ is right-continuous if all $f_n$
 % are right-continuous, and also that $F_{N+1} = F_N + f_{N+1}$ is
 % right-continuous. As this applies for all $N$, it follows from
 % induction that $\lim_N F_N$ is right-continuous, provided this
 % limit exists. When $f_n$ are all increasing, which is the case for
 % our situation by taking $f_n(t) = 1\{A_n \leq t\}$, then this
 % limit certainly exists.
\end{solution}
\end{extra}


\opt{solutionfiles}{
\Closesolutionfile{hint}
\Closesolutionfile{ans}
\subsection*{Hints}
\input{hint}
\subsection*{Solutions}
\input{ans}
}
%\clearpage

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../companion"
%%% End:
