\section{Renewal Reward Theorem}
\label{sec:renew-reward-theor}


We state the \recall{renewal reward theorem} and provide graphical motivation for its validity.
This results proves to be extremely useful: we will apply it regularly in the sequel of the book and here we use it to provide another definition of the load~$\rho$. 


\opt{solutionfiles}{
\subsection*{Theory and Exercises}
\Opensolutionfile{hint}
\Opensolutionfile{ans}
}


In essence the renewal reward theorem is very simple: it states that when customers arrive at rate $\lambda$ and each customer pays an average amount $X$, the system earns money at rate $Y=\lambda X$.
\Cref{fig:renewal} provides graphical motivation about why this theorem is true; \citet{el-taha98:_sampl_path_analy_queuein_system} gives a (simple) proof.

\begin{theorem}[Renewal Reward Theorem, $Y=\lambda X$]
 Consider epochs $\{T_k, k=0, 1, \ldots\}$ such that $0=T_0 < T_1 < \cdots$
 Let $N=\{N(t), t\geq 0\}$ be the associated counting process with $N(t) = \max\{k : T_k \leq t\}$.
 Let $\{Y(t), t\geq 0\}$ be a non-decreasing right-continuous (deterministic) process.
 Define $X_k = Y(T_k)-Y(T_{k-1})$.
 Suppose that $N(t)/t\to\lambda$ as $t\to\infty$, where $0<\lambda < \infty$.
 Then $Y(t)/t$ has a limit iff $n^{-1}\sum_{k=1}^n X_k$ has a limit, and then $Y=\lambda X$. In other words, 
 \begin{equation*}
 \lim_{t \to \infty} \frac{Y(t)}t=Y \iff \lim_{n \to \infty} \frac 1n\sum_{k=1}^n X_k = X, 
 \end{equation*}
and then $Y=\lambda X$. 
\end{theorem}


\begin{figure}[ht]
 \centering
\begin{tikzpicture}[scale=1]
%axis
\draw[->] (0,0) -- coordinate (x axis mid) (8.5,0);
\draw[->] (0,0) -- coordinate (y axis mid) (0,5.5);
%\node[below=0.2cm] at (x axis mid) {$t$};

\draw plot [smooth] coordinates {(1,0.7) (2,1) (3,1.8) (4,2) (5,3) (6,3.2) (7, 3.4) (7.5,4.0)};
\node[right] at (7.1,3.4) {$Y(t)$};
\node at (7.,-0.5) {$t$};

\node at (2.,-0.5) {$T_{k-1}$};
\draw[dotted] (2,0)--(2,1);
\draw[dotted] (2,1)--(3,1);


\node at (3.,-0.5) {$T_{k}$};
\draw[dotted] (3,0)--(3,1.8);
\draw[dotted] (3,1.8)--(5,1.8);

\draw [
 thick,
 decoration={brace, mirror, raise=0.1cm },
 decorate
] (3,1) -- (3,1.8)
node[pos=0.5, xshift=0.6cm] {$X_k$}; 


\node at (5.,-0.5) {$T_{k+1}$};
\draw[dotted] (5,0)--(5,3);

\draw [
 thick,
 decoration={brace, mirror, raise=0.1cm },
 decorate
] (5,1.8) -- (5,3)
node[pos=0.5, xshift=0.7cm] {$X_{k+1}$}; 

\node[right] at (0.5,4) {$\frac{Y(t)}t \approx \frac{\sum_i^n X_i}{T_n} = \frac{n}{T_n} \frac{\sum_i^n X_i}n $};


\draw[dotted,<->, =stealth] (7,0.05)--(7,3.35) node[midway, rotate=90, fill=white] {$\sum_{i=1}^n X_i$};
\draw[dotted,<->, =stealth] (0.05,0.5)--(6.95,0.5) node[pos=0.85, fill=white] {$T_n$};
\end{tikzpicture}
\caption{A graphical `proof' of $Y=\lambda X$. Here $Y(t)/t\to Y$, $n/T_n\to \lambda$ and $n^{-1}\sum_i^n X_i \to X$. Observe that in the
 figure $X_k$ does not represent an inter-arrival time; instead it
 corresponds to the increment of (the graph of) $Y(t)$ between two
 consecutive epochs $T_{k-1}$ and $T_k$ at which $Y(t)$ is
 observed.}
 \label{fig:renewal}
\end{figure}

Let us use this theorem to understand the idea of `load' in a different way. 
Define the \recall{load} or \recall{utilization} as the limiting fraction of time the server is busy, i.e.,
\begin{equation*}
 \rho = \lim_{t\to\infty} \frac 1 t \int_0^t \1{L(s)>0} \d s.
\end{equation*}

\begin{exercise}\clabel{ex:l-162}
Use the renewal reward theorem to prove that $\rho = \lambda \E S$ for the rate-stable $G/G/1$ queue. 
\begin{hint}
 Define $Y(t) = \int_0^t \1{L(s)>0} \d s$ as the total amount of time the server has been busy up to the time $t$.
 Then take as epochs $T_k = D_k$ and use rate stability.
\end{hint}
\begin{solution}
 It is evident that $X_k = Y(D_k)-Y(D_{k-1})=S_k$, hence $X = \lim_{n\to\infty} n^{-1}\sum_{k=1}^n X_k = \E S$.
 Also $\lim_{t\to\infty} Y(t)/t=\rho$.
 Finally, in the relation $Y = \lambda X$, the $\lambda$ is $\delta$ since we consider departure epochs $T_k = D_k$, rather than $A_k$.
 By the renewal reward theorem $Y=\lambda X$ we get that $\rho = \delta \E S$.
 Finally, by rate-stability, the job arrival rate $\lambda = \delta$, hence $\rho = \lambda \E S$.
\end{solution}
\end{exercise}


\begin{exercise}\clabel{ex:l-163}
 We can derive the relation $\rho = \lambda \E S$ in a somewhat more direct way by considering the fact that
\begin{equation*}
 \sum_{k=1}^{A(t)} S_k \geq \int_0^t \1{L(s)>0} \d s \geq \sum_{k=1}^{D(t)} S_k.
\end{equation*}
Explain this, and complete the argument.
\begin{solution}
Observe that
since $t$ can lie half way a service interval and $A(t) \geq D(t)$. As
$A(t)\to \infty$ as $t\to\infty$,
\begin{equation*}
 \lim_{t\to\infty} \frac 1 t\sum_{k=1}^{A(t)} S_k = 
 \lim_{t\to\infty} \frac{A(t)}t \frac{1}{A(t)} \sum_{k=1}^{A(t)} S_k = 
 \lim_{t\to\infty} \frac{A(t)}t \cdot \lim_{t\to\infty}\frac{1}{A(t)} \sum_{k=1}^{A(t)} S_k = \lambda \E S.
\end{equation*}
Applying similar limits to the other inequality gives
\begin{equation*}
\lambda \E S \geq \rho \geq \delta \E S.
\end{equation*}
Hence, if the system is rate-stable, $\delta=\lambda$, and then $\rho = \lambda \E S$. 

Note that this is in fact the same argument as that underlies the renewal reward theorem. Henceforth we will just use the renewal reward theorem. 
\end{solution}
\end{exercise}



\opt{solutionfiles}{
\Closesolutionfile{hint}
\Closesolutionfile{ans}
\subsection*{Hints}
\input{hint}
\subsection*{Solutions}
\input{ans}
}
%\clearpage


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../companion"
%%% End:
